---
title: "Final Project"
author: "Kyle McGrath, Elias Wilson, Benjamin Fick"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Section 1

We choose to conduct research on the relationship between a country's access to a close reliable water source, and the gross domestic product per capita. We hypothesize that countries with access to reliable water sources would have a higher GDP than countries without access to basic water. 


**Setup**

```{r setup}
library(tidyverse)
library(lubridate)
library(broom)
library(kableExtra)

# fix the file path for your file folder structure
gdp <- read_csv("gdp_pcap.csv", show_col_types =FALSE)
water <- read_csv("at_least_basic_water_source_overall_access_percent.csv", show_col_types = FALSE)
```

# Data Description and Variable Description

The At least Basic Water Source, overall access (%) dataset contains values representing the percentage of a country's population who have access to some sort of modified water source (like a well) within 30 minutes travel of them, with values for each year from 2000 to 2020. Values range from 0 - 100 representing the percentage of basic water source access. If values are missing they are expressed as NA. 

**Variables**

* country - A string containing the country 
* 2000:2020 - A double ranging from 0-100 containing the basic water source overall access percent for that particular year

The GDP per capita (price and inflation adjusted, in PPP $2017) dataset contains values representing the gross domestic product of a country divided by its population, with values for every year. The values obtained from 1990 to 2019 were obtained from the GDP per capita data from the World Bank, which is the data we will use in our study. Other values were obtained from the Maddison Project Database or the Penn World and then combined into this larger dataset. The dataset continues well into the future from the year 1800 to 2100 based on projections. There are no missing values as missing values were input using projections. 

**Variables**

* country - A string containing the country 
* 1800:2100 - A double containing the gross domestic product for that year or a string written in the format “1.1k” representing the gross domestic product for that year. 


# Pivot Longer

For our study we chose to only analyze the years 2000 to 2020 as this was the extent of the At least Basic Water Source dataset. For both datasets we used the pivot_longer function selecting the years ‘2000’:’2020’ as our columns. 

```{r}
#Pivot
gdp_pivoted <- gdp |>
  pivot_longer(cols = '2000':'2020', 
               names_to = "year",
               values_to = "gdp") |>
  select(country, year, gdp)

water_pivoted <- water |>
  pivot_longer(cols = '2000':'2020',
               names_to = "year",
               values_to = "water")
```


# Join the datasets

The At least Basic Water Source dataset had less countries observed in their study. For this reason when joined, we used an inner join as we only wanted to keep rows in both tables. We set join_by to the country and the year. 

```{r}
#Join
joined <- inner_join(water_pivoted, gdp_pivoted, join_by(country, year))
```

# Cleaning

For convenience, we chose to clean the data after joining. To clean the data we had to convert some GDP values from the format "123.4k" to integers of the format 123,400. To do this we mutated the joined dataset using and ifelse statement to detect values ending with “k”, remove these values, and multiply them by 1000. The missing values from the water source dataset were imported as NA’s, so we didn’t have to clean these. The GDP dataset didn’t have any NA’s, so this wasn’t an issue either. 


```{r filtering out ks}
joined_clean <- joined |>
  mutate(gdp = ifelse(str_detect(gdp, "k$"), as.numeric(str_remove(gdp, "k$")) * 1000,
                      gdp),
         gdp = as.numeric(gdp),
         water = as.numeric(water)) |>
  mutate_all(~replace_na(., 0))
```
# Conclusion

We hypothesize that any increase in access to water will have a corresponding increase in GDP, either immediately or shortly following the change. We will explore this claim with a linear regression and simulation. 

# Section 2

## Section 2.1

# Visualizing Relationship Between our Variables

```{r}
joined_clean |>
  group_by(country) |>
  summarize(avg_water = mean(water), avg_gdp = mean(gdp)) |>
  ggplot(aes(x = avg_water, y = avg_gdp)) +
  geom_point() 
```

Here is the initial scatterplot with the fitted regression model. From the graph, we can see that the relationship between average percent access to water and average GDP is quite exponential. Countries with 100% access to water on average have a much higher average GDP. However, it seems that countries with anything less than 80% access to water on average have roughly the same GDP on average. It also seems as though there is an outlier present, as one country has less than 20% access to water on average. Also, it seems as though taking the average of both percent access to water and GDP make it much easier to visualize, so this is what we will be doing going forward.

# Visualizing Relationship Over Time

```{r}
joined_clean |>
  ggplot(aes(x = water, y = gdp)) +
  geom_point() +
  facet_wrap(~year)
```

Looking at the relationship over time, we can see that is hasn't changed much. There is an obvious exponential trend in the data between our percent access to water variable and the overall GDP. This is helpful, because we can see now that in order for our linear regression model in the next step to make sense, we will need to perform a transformation or two. 

# Visual Linear Regression

```{r}
joined_clean |>
  group_by(country) |>
  summarize(avg_water = mean(water), avg_gdp = mean(gdp)) |>
  ggplot(aes(x = avg_water, y = avg_gdp)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Here is our linear regression model before performing any transformations. The linear model overlaid above attempts to predict values of average GDP for each value of average access to water. The model uses the given data from our cleaned data set, and the grey portion above and below the line represent the standard error of the model. It doesn't fit our data very well, but the relationship we observed between our two variables previously explains this.

```{r}
joined_clean2 <- joined_clean |>
  group_by(country) |>
  summarize(avg_water = mean(water), avg_gdp = mean(gdp))

joined_clean2 |>
  ggplot(aes(x = avg_water, y = log(avg_gdp))) +
  geom_point() +
  geom_smooth(method = "lm")
```

Here is our linear regression model after transforming our response variable. Due to the positive exponential relationship between our two variables, it made sense to take the natural log of average GDP. Also, our model now reflects a more accurate linear relationship between average percent access to water and average GDP.

# Calculate Regression Equation

```{r}
initial_model <- lm(log(avg_gdp) ~ avg_water, data = joined_clean2)
initial_model
```

Here is our estimated regression equation: log(yhat) = 0.055(avg_water) + 4.583 where yhat is predicted average GDP.

# Interpreting the Regression Coefficients

Intercept: 5.196 is the predicted log(avg_gdp) when all of the predictors are 0 and at reference level.  We can also say that e^5.196 or 180.55 is the predicted average GDP when all of the predictors are zero. In this case, we only have one predictor, so this is irrelevant. However, our intercept overall is also quite irrelevant, as the interpreted value is extrapolated. There is no data available for countries with no access to water.

Slope: Our model predicts that for a one percent increase in average access to water, the average GDP increases by e^0.048 or 1.05. 

# Looking at our Model Fit

```{r}
aug_model <- augment(initial_model) 

colnames(aug_model) <- c("Response", "Water", "FittedValues", "Residuals", "Hat", "Sigma", "Cooked", "StdResid")

tab2 <- aug_model |>
  summarize(response_var = var(Response), 
            response_fitted = var(FittedValues),
            response_resid = var(Residuals))

colnames(tab2) <- c("Response", "Fitted Values", "Residuals")

kable(tab2, align = "ccc", format = "html") |>
  add_header_above(c("Variance of Specific Values" = 3))
```

We can see above that the average variance in our response variable is the higher, where as our variance in the residuals is the highest. This means that our model does a fairly good job of predicting average GDP.

# Exploring "Model Quality"

```{r}
expl_var <- tab2[2]/tab2[1]

kable(expl_var, align = "c", format = "html", col.names = c("R-Squared"))
```

Based on the R^2 value above, we can see that 57.15% of the variability in average GDP is explained by our model based on average percent access to water. This means that our model does a fairly good job of predicting average GDP.
